NURM: a tool to locate numerical differences in pipelines

### 2019-09-19 paper.tex
Add Results and Discussion section

In particular, the effect of the operating system is quantified 
in~\cite{Glatard2015, Gronenschild2012} specially on some of the main 
neuroimaging pipelines including FSL, CIVET and FreeSurfer pipelines. 
The binary differences are measured by the checksum method between 
results of the same analyses on different operating systems, which 
indicate a significant disparity. 
}

@@

The variety of computing infrastructures, particularly the operating system, are 
known to influence the reproducibility of computational pipelines because of the
creation of small numerical differences. 
These differences mainly correspond to the mathematical functions implemented 
in different operating system libraries.
For instance, changing the mathematical functions like \emph{expf()} and 
\emph{cosf()} which manipulate the precision of floating-point representations, 
between \emph{glibc} libraries in different operating systems can produce 
small numerical differences.
Furthermore, a similar issue is expected for any operating system which is 
based on \emph{glibc}, the GNU C library.

Indeed, the propagation and amplification of these tiny perturbations by analysis 
may cause reproducibility issues. We focus on the differences 
created by different version of operating systems in computational pipelines. 
In particular, NURM-tool is tested on the neuroimaging applications which are 
predominantly using mathematical libraries.

\subsection{HCP pipelines and datasets}

@@

According to~\cite{glasser2013}, main steps of the HCP PreFreeSurfer pipeline are:
(1) Distortion Correction: 
is the correction of geometric distortions for routine MRI scans 
to produce an undistorted native structural volume space for each subject.
(2) Anatomical Average: average same modality images (if more than one is available)
(3) ACPC (Anterior Commissure, Posterior Commissure) Alignment: 
is aligning images to the MNI space to have the same orientation as 
the reference template.
(4) Brain Extraction: includes a robust initial brain extraction, using FSL's linear (FLIRT) 
and non-linear (FNIRT) registration. 
(5) Bias Field Correction: applies intensity inhomogenity correction on the registered images.
Calculates bias field using square root of the product of T1w and T2w images. 
(6) Atlas-Registration: register the subject's native structural volume space to MNI space as the last step.

@@

A detailed description of the analyses done in the
pipelines is available in~\cite{glasser2013}.

HCP data consists of brain images collected from an individual healthy person, 
in the age range from 22-35 years~\cite{van2013wu}. Data from 1200 subjects 
is available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
Such a directory containing the HCP data is denoted by the term ``subject".
In this paper, we randomly selected 100 unprocessed subjects from 
the HCP data release S500. All the subjects were scanned using the 3-tesla 
type of MRI scanners.

@@

It should be noted that some files may have the same data but 
different metadata in headers, including creation time, modification time and date, 
which turns to non-identical checksums. 
So, we added some other functions for the file comparison to be sure that differences 
are in part of the file data. 
Depending on the file type, we used functions from FreeSurfer software package to determine 
whether two files differ, including  
\emph{mri\_diff}, \emph{mris\_diff}, and \emph{lta\_diff} for comparing volumes, surface-based files, 
and lta files (transforms) respectively. Also, text files are compared using a custom function.

@@

In our experiments, we identified the processes that create differences due to 
using random variables in the FreeSurfer pipeline. 
For example, process \emph{mris\_curvature}
computes mean and Gaussian curvatures of a cortical surface using the random variable.
To avoid such randomness we 
initialized random number generators using a seed state in 
the command line in which pipeline script calls the program.
This ensures that results are the same from run to run.
Therefore, NURM-tool can identify the process in the pipelines 
that create differences caused by the inter-OS differences.

@@

Figure~\ref{fig:fnirt_result} shows differences in the \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz} files. 
These files are created as a part of ACPC-Alignment (Step 3 of PreFreeSurfer).
The RMSE (root mean squared error) value of results is 0.029 and the dice coefficient is 0.125.
The square pixels in the brain in this checkerboard images show the differences between results in two OSes. 
In the next subsection, we will show that such misalignment is created due to 
non-linear registration in \emph{FNIRT} process in PreFreeSurfer pipeline.

Furthermore, Figure 
\ref{fig:tissue_class} illustrates the binarized differences in the \emph{aseg.hires.mgz} files.
This is the result of brain tissue segmentation across OSes in FreeSurfer pipeline. 
Each color represents a different structure in the brain.
The bright spots show relevant pixels in the image which are different.
The RMSE and the dice coefficient values of results are 0.009 and 0.991 respectively.

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/segmentation.png} 
    \caption{Differences between FNIRT results from PreFreeSurfer 
    (checkerboard image \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz}) (CentOS6 vs. 
    CentOS7). Right images show the zoomed version part of the images marked 
    by the red rectangle.} 
    \label{fig:fnirt_result}
\end{figure}

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_classification.png} 
  \caption{Binarized differences between brain segmentation results from 
           FreeSurfer (CentOS6 vs. CentOS7).} 
  \label{fig:tissue_class}
\end{figure}

@@

First, subjects are clustered into different groups using the clustering method as described before.
We choosed the threshold value of zero for the clustering method. 
This turns to have the subjects with the same topology in each cluster.
Also, we used the nearest neighbor algorithm to calculate the distance between clusters.
Table~\ref{table:data-clusters} shows the PreFreeSurfer result of subject clustering. 
Among our dataset, we identified 4 types of subjects with 
different topology of \emph{process trees}. 
Differences correspond to the data processing that leads to producing disparate output file(s).
For example, subject type 3 missed output file \emph{T1w2\_gdc.nii.gz} after processing 
in comparison with subject type 1.
Besides, we verified that 
the \emph{process trees} were identical for all subjects of the same type in 
different versions of CentOS.

@@

Then, we determined the frequency of the processes that create differences in 
the pipeline in Figure~\ref{fig:pfs_freq}. \emph{FLIRT} is the most frequent 
process that produces differences.
Furthermore, the percentage of processes that introduce differences 
in PreFreeSurfer along with the 
pipeline elements are depicted in Figure~\ref{fig:pfs_freq} 
including linear registration with “\emph{FLIRT}” (in 
ACPC-Alignment, BrainExtraction, DistortionCorrection, 
AtlasRegistration), non-linear registration with “\emph{FNIRT}” (in 
BrainExtraction and AltasRegistration), image warping with 
“\emph{new\_invwarp}” (in BrainExtraction and AtlasRegistration). 
Also, differences were observed in image mean and standard-deviation 
computations in image interpolation with “\emph{fslmaths}” (in AnatomicalAverage). 
It can be seen that \emph{fslmaths} and \emph{FLIRT} 
are the processes that at least have one subprocess to produce differences across all the subjects.

@@

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/pfs_heatmap.png}
  \caption{Colorized processes along with the pipeline elements that introduce 
           differences among 20 subjects. The range between red and green colors show the most and 
           least frequency of processes respectively. Besides, the number 
           of occurrences of each processes with differences is represented in the line below the heatmap.}
  \label{fig:pfs_freq}
\end{figure}

@@

We found that inter-OS differences occur if pre-processed data are different in FreeSurfer. 
Figure~\ref{fig:tissue_class} shows differences that coming from PreFreeSurfer analysis 
as a result of propagation of differences. 
NURM tool identified that there are no processs in the pipeline that create differences.
This is because of using static builds of FreeSurfer which statically linked libraries (embeded libraries in the pipeline).

@@

\note{Pipelines amplify small numerical differences because they are numerically 
unstable. Furthermore, math libraries evolve over time, leading to 
different numerical errors. we listed some of the irreproducibility 
causes of the pipelines as we mentioned in the previous section 
overally along with exact command arguments.

Mention that this was only possible because the unprocessed data was 
shared in the first place. DICOM to Nifti conversion was out of scope 
and may introduce other issues.}

@@

\note{Our technique is able to characterize the stability of a pipeline's 
components automatically. The numerical instability in the 
PreFreeSurfer HCP pipeline arises mainly from linear and non-linear 
registration processes implemented in FSL FLIRT and FNIRT.

There are a few ways to impede such instabilities:
\begin{itemize}
\item Use a single operating system
\item Containerize pipelines
\item Increase numerical precision
\item Be stricter on truncation and rounding standards (IEEE 754)
\item Build static executable
\end{itemize}

The results still suffer from small perturbations literally because of 
the fact that pipeline are not numerically stable. The preferred 
solution is to detect and fix numerical instability of the pipeline 
instead of masking the problem. These processes need to be reviewed to 
understand and correct the cause of instabilities. 
}

@@

@@

\note{CBRAIN team. Compute Canada(Calcul Quebec).}

>>>>

In particular, the effect of 
the operating system across some of the main 
neuroimaging pipelines including FSL, CIVET and FreeSurfer pipelines 
is studied in~\cite{Glatard2015, Gronenschild2012} that indicates a significant disparity. 
}

@@

The variety of computing infrastructures, particularly the operating system, is 
known to influence the reproducibility of computational pipelines because of the
creation of small numerical differences. 
These differences mainly correspond to the mathematical functions implemented 
in different operating system libraries.
For instance, changing the mathematical functions like \emph{expf()} and 
\emph{cosf()} which manipulate the precision of floating-point representations, 
between \emph{glibc} libraries in different operating systems can produce 
small numerical differences.
Besides, a similar issue is expected for any operating system which is 
based on \emph{glibc}, the GNU C library.

Indeed, the propagation and amplification of these tiny perturbations by analysis 
may cause a significant reproducibility issue. We focus on the differences 
created by different version of operating systems in computational pipelines. 
In particular, NURM-tool is tested on the neuroimaging applications which are 
predominantly using mathematical libraries.

\subsection{HCP pipelines and dataset}

@@

According to~\cite{glasser2013}, main sub-pipelines of the HCP PreFreeSurfer are:
(1) Distortion Correction (DC) 
to correct the geometric distortions for routine MRI scans 
and produce an undistorted native structural volume space for each subject,
(2) Anatomical Average (AAve) to average same modality images (if more than one is available),
(3) ACPC (Anterior Commissure, Posterior Commissure) Alignment (ACPC-A) 
is aligning images to the MNI space to have the same orientation as 
the reference template,
(4) Brain Extraction (BExt) includes a robust initial brain extraction, using FSL's linear (FLIRT) 
and non-linear (FNIRT) registration, 
(5) Bias Field Correction (BFC) applies intensity inhomogenity correction on the registered images, 
%Calculates bias field using square root of the product of T1w and T2w images. 
(6) Atlas-Registration (AR) to register the subject's native structural volume space to MNI space as the last step.

@@

A detailed description of the analyses done in these
pipelines is available in~\cite{glasser2013}.

HCP data consists of brain images collected from an individual healthy person, 
in the age range between 22-35 years~\cite{van2013wu}. Data from 1200 subjects 
is available in \href{https://db.humanconnectome.org}{the ConnectomDB repository}. 
Such a directory containing the HCP data is denoted by the term ``subject".
In this paper, we randomly selected 20 unprocessed subjects from 
the HCP data release S500. All the subjects were scanned using the 3-tesla 
type of MRI scanners.

@@

It should be noted that some files may have the same data but 
different metadata in headers, including creation time, modification time and date, 
which turns to non-identical checksums. 
So, we added some other functions for the file comparison to be sure that differences 
are in part of the file data. 
Depending on the file type, we used functions from FreeSurfer software package to determine 
whether two files differ, including  
\emph{mri\_diff}, \emph{mris\_diff}, and \emph{lta\_diff} for comparing volumes, surface-based files, 
and lta files (transforms) respectively. 
These functions check data based on the volume dimentions, 
resolutions, pixel intensities,  
the number of surface vertices and neighbors, and geometry information. 
It then returns files differ if there be differences in any of these parameters.
Also, text files are compared using a custom function that compares the content line by line.

@@

In our experiments, we identified the processes that create differences due to 
intra-OS differences.  
For example, process \emph{mris\_curvature}
computes mean and Gaussian curvatures of a cortical surface using 
the random variable in FreeSurfer.
To avoid such randomness we 
initialized random number generators using a seed state in 
the command line in which pipeline script calls the program.
This ensures that results are the same from run to run.
Therefore, NURM-tool can identify the processes in the pipelines 
that create inter-OS differences.

@@

Figure~\ref{fig:fnirt_result} shows differences in the files 
that are created as a part of ACPC-Alignment in PreFreeSurfer.
The square pixels in the brain in the checkerboard images show the differences between results in two OSes. 
The RMSE (root mean squared error) value of results is 0.029 and the dice coefficient is 0.125.
In the next subsection, it is shown that such misalignment is created due to 
non-linear registration in \emph{FNIRT} process in PreFreeSurfer pipeline.

Furthermore, Figure 
\ref{fig:tissue_class} illustrates the heatmap binarized differences 
of the brain tissue segmentation across OSes in FreeSurfer pipeline.
These differences are accumulated from 20 subject results.
So, the magnitude of differences is in the range of 1-20 as the total number of subjects.
The darker blue pixels show differences among less subjests and 
the brighter blue pixels are related to the differences among more subjects.

%The brighter spots show relevant pixels in the segmented image which are different among more subjects.
%The RMSE and the dice coefficient values of results are 0.009 and 0.991 respectively.
%Each color represents a different structure in the brain.

\begin{figure}
  \centering
    \includegraphics[width=\columnwidth]{images/segmentation.png} 
    \caption{Differences between FNIRT results of one subject from PreFreeSurfer 
    (checkerboard image \emph{T2w\_acpc\_to\_MNI\_nonli.nii.gz}) (CentOS6 vs. 
    CentOS7). Right images show the zoomed version part of the images marked 
    by the red rectangle.} 
    \label{fig:fnirt_result}
\end{figure}

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{images/brain_classification.png} 
  \caption{Heatmap binarized differences between brain segmentation results (file \emph{aseg.hires.mgz}) from 
          20 subjects in FreeSurfer (CentOS6 vs. CentOS7).} 
  \label{fig:tissue_class}
\end{figure}

@@

First, subjects are clustered into different groups using the clustering method as described before.
We set the threshold value to zero for the clustering method. 
This turns to have the subjects with the same topology in each cluster.
Also, we used the nearest neighbor algorithm to calculate the distance between clusters.
Table~\ref{table:data-clusters} shows the PreFreeSurfer result of subject clustering. 
Among our dataset, we identified 4 types of subjects with 
different topology of \emph{process trees}. 
Differences correspond to the data processing that leads to producing disparate output file(s).
For example, subject type 3 missed output file \emph{T1w2\_gdc.nii.gz} after processing 
in comparison with subject type 1.
Besides, we verified that 
the \emph{process trees} were identical for all subjects of the same type in 
different versions of CentOS.

@@

Then, we determined the frequency of the processes that create differences in 
the pipeline in Figure~\ref{fig:pfs_freq}. \emph{FLIRT} is the most frequent 
process that produces differences.
Furthermore, the percentage of processes that introduce differences 
along with the 
sub-pipelines are depicted in this figure 
including linear registration with “\emph{FLIRT}” (in 
ACPC-Alignment, Brain Extraction, Distortion Correction, 
Atlas Registration), non-linear registration with “\emph{FNIRT}” (in 
Brain Extraction and Altas Registration), image warping with 
“\emph{new\_invwarp}” (in Brain Extraction and Atlas Registration). 
Also, differences were observed in image mean and standard-deviation 
computations in image interpolation with “\emph{fslmaths}” (in Anatomical Average). 
It can be seen that \emph{fslmaths} and \emph{FLIRT} 
are the processes that at least have one subprocess to produce differences across all the subjects.

@@

\begin{figure}
\centering
  \includegraphics[width=\columnwidth]{images/pfs_heatmap.png}
  \caption{Heatmap of difference frequency of processes along with the sub-pipelines that introduce 
           differences among 20 subjects. The range between red and green colors show the most and 
           least frequency of processes respectively. Besides, the total number 
           of occurrences of each processes is represented in the bracket in each cell.}
  \label{fig:pfs_freq}
\end{figure}

@@

We executed FreeSurfer pipeline on both OSes using a fixed input data, result of PreFreeSurfer on Centos 6, 
to determine the creation of differences. 
NURM tool identidied \emph{mris\_make\_surfaces} as the only process in the pipeline that create differences.
This process generates surface files for cortical and white matter surfaces, curvature file for cortical thickness and 
surface file estimate for layer IV of cortical sheet. 
In this analysis, the number of process that create differences are dropped becasue of 
using the static builds of FreeSurfer in which libaries are embeded in the pipeline package.

However, we found that inter-OS differences occur when pre-processed data are different in FreeSurfer. 
Using the result data from PreFreeSurfer, we executed FreeSurfer pipeline in each operating system separately.
FreeSurfer pipeline is unstable as it propagates and amplifies differences 
through the pipeline analysis.
Figure~\ref{fig:tissue_class} shows inter-OS differences  
as a result of propagation and amplification of differences. % that coming from PreFreeSurfer outputs.

@@

We indentified 4 processes that create differences 
among all 28 distinct processes that are involved in the PreFreeSurfer execution. 
The numerical instability in the 
PreFreeSurfer HCP pipeline arises mainly from linear 
registration processes implemented in FSL \emph{FLIRT}. 
We can see that \emph{FLIRT} creates differences in almost all the sub-pipelines. 
Therefore, it is important to more investigate about the registration procedure and find 
a way to stabilize this process.

In the second experiment, 
the analysis results obtained by the pipeline show that inter-OS differences are visually substabtial.
We see that statically builds of the pipeline improves reproducibility, but 
it would not improve the numerical stability of the pipeline. 
FreeSurfer pipeline propagates and amplifies the numerical differences that are created by the 
differences in the underlying operating system libraries.

In addition to variations of the operating system, there are other factors that may affect 
the results such as software versions, compilers, hardwares, etc.
Generally, we can use the Monte Carlo Arithmatic (MCA) to determine the numerical instability of the pipelines.
This argorithm perturbs every floating-point operation 
in the analysis based on the random rounding in each run. 
By this algorithm, NURM tool can easily locates which parts of the analysed source code are to be responsible 
for floating-point related instabilities.

Although we could identify the processes in the pipeline that create differeces, the functions 
that amplify these differences should be quantified in the future works.
For this purpose, we can evaluate the final result after perturbing the identified processes in the pipeline iterratively.
As a result, the processes that amplify differences would have a significant effect on the results independently.

@@

Our technique is able to characterize the stability of a pipeline
automatically. We identified the main causes of the inter-OS differences as (1) the evolution 
of the math libraries over the time and, (2) the instability of the pipelines. 
There are two ways to tackle this problem. The easy but less preferred solution is masking the instabilities.
This can be done by using, (1) single operating system for the processing of subjects, 
(2) containerizing the pipelines so that the 
processing is done on a more controlled environment, 
(3) increasing the numerical precision of the arithmetics, 
(4) building static executables by removing the host operating system library dependencies. 
These solutions only makes the problem invisible but the instability still remains.

The preferred solution is to find the origin of amplification of the differences and 
then fixing the particular functions/modules in the processing pipelines.

@@

\note{
CBRAIN team. Compute Canada(Calcul Quebec).

@@

}
